---
title: "esm-206-assignment-5-q3"
author: "Ken Gilliland"
date: "November 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}

# load necessary packages

library(tidyverse)
library(pwr)
library(knitr)
library(kableExtra)
library(plotly)
library(ggrepel)
library(effsize)
library(pwr)
library(vcdExtra)
library(car)



```

```{r, include= FALSE}
# Load necessary data

phd_salary <- read_csv('Median salary for doctoral recipients.csv')

# Reformat phd_salary dataframe

phd_salary_employment <- phd_salary %>% 
  select(field, employment.male, employment.female) %>% 
  rename(male = employment.male,
         female = employment.female) %>% 
  gather(key = sex, value = employment, 2:3)

phd_salary_postdoc <- phd_salary %>% 
  select(field, postdoc.male, postdoc.female) %>%
  rename(male = postdoc.male,
         female = postdoc.female) %>% 
  gather(key = sex, value = postdoc, 2:3)

phd_salary2 <- full_join(phd_salary_employment, phd_salary_postdoc,
                         by = c('field', 'sex'))

as.vector(phd_salary2) %>% 
  filter(employment != -99999) %>% 
  filter(postdoc != -99999) # filtering out NAs  

View (phd_salary2)
```


```{r}
#histogram for employment

median_hist_emp<- ggplot(phd_salary2, aes(x = employment)) +
  geom_histogram(aes(fill = employment)) +
  facet_wrap(~ sex, scale = "free")

median_hist_emp
```

```{r}
qq_employment<- ggplot(phd_salary2, (aes(sample= employment))) +
  geom_qq(aes(color = sex)) +
  facet_wrap(~sex, scale = "free")

qq_employment
```

```{r}
#histogram for postdoc

median_hist_post<- ggplot(phd_salary2, aes(x = postdoc)) +
  geom_histogram(aes(fill = postdoc)) +
  facet_wrap(~ sex, scale = "free")

median_hist_post
```

```{r}
qq_postdoc<- ggplot(phd_salary2, (aes(sample= postdoc))) +
  geom_qq(aes(color = sex)) +
  facet_wrap(~sex, scale = "free")

qq_postdoc
```

based on qqplots and histograms observations for each group appear to not be normally distrubuted. Also, n < 15 for all groups, so means may not be distrubuted

Perform a Mann-Whitney U

Professional male and female salaries

```{r}
employment_sal <- phd_salary2  %>% 
    select(employment, sex) # keeps postdoc and sex

View (employment_sal)
```

```{r}
#Is there a significant difference in employment salary for males and females?

# Unpaired: Mann-Whitney U

mwu_employment <- employment_sal %>%
  wilcox.test(employment ~ sex, data = .)

mwu_employment

# There is no significant difference in employment salaries between males and females.
```

A Mann-Whitney U test revealed that there was not a significant difference between rankings of professional salaries between male and female doctorate recipients in 2015 (W = 88.5, p = 0.32). 

```{r}

sex_names <- c("Female", "Male")

sex_color <- c("white", "gray60")

employment_box <- ggplot(employment_sal, aes(x = sex, y = employment)) +
  geom_boxplot(aes(fill = sex)) +
  scale_fill_manual(values = sex_color) +
  theme_classic() +
  scale_x_discrete(labels = sex_names) +
  labs(x = "Sex", y = "Median Professional Salary (dollars) (n = 30)") +
  theme(plot.title = element_text(hjust = 0.5))
  

employment_box
```
Postdoc male and female salaries

```{r}
postdoc_sal <- phd_salary2  %>% 
    select(postdoc, sex) # keeps postdoc and sex

View (postdoc_sal)
```

```{r}
#Is there a significant difference in postdoc salary for males and females?

# Unpaired: Mann-Whitney U

mwu_postdoc <- postdoc_sal %>%
  wilcox.test(postdoc ~ sex, data = .)

mwu_postdoc

# There is no significant difference in postdoc salaries between males and females.

```

A Mann-Whitney U test revealed that there was not a significant difference between rankings of postdoc salaries between male and female doctorate recipients in 2015 (W = 108, p = 0.87).  



```{r}

sex_names <- c("Female", "Male")

sex_color <- c("white", "gray60")

postdoc_box <- ggplot(postdoc_sal, aes(x = sex, y = postdoc)) +
  geom_boxplot(aes(fill = sex)) +
  scale_fill_manual(values = sex_color) +
  theme_classic() +
  scale_x_discrete(labels = sex_names) +
  labs(x = "Sex", y = "Median Postdoctoral Salary (dollars) (n = 30)") +
  theme(plot.title = element_text(hjust = 0.5))
  

postdoc_box
```



