---
title: "esm-206-assignment-5-q2"
author: "Kate"
date: "November 29, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Q2: Shifts in female PhD recipients by field (1985, 2000, and 2015)

Describe if and how there was a shift in PhDs awarded to females in four fields (Physical and Earth Sciences, Engineering, Education, and Humanities & Arts) in 1985, 2000, and 2015. Describe your results statistically, in a graph or table, and in text. Note: There are several ways that you can interpret this question. You are invited to decide which you think is/are most interesting. Just be really clear about what you are asking/answering in your report.

```{r packages & data, include=FALSE}

# Load necessary packages

library(tidyverse)
library(RColorBrewer)

# Load necessary data

phds_csv <- read_csv('PhDs by Field 1985 - 2015.csv')

```


```{r reformatting dataframes, include=FALSE}

# Reformatted, simplified, truncated dataframe

phds <- phds_csv %>% 
  gather(key = year, value = doctorates, 3:9) %>% 
  filter(sex != 'total',
         year == 1985 | year == 2000 | year == 2015,
         field == 'Physical sciences and earth sciences' |
           field == 'Engineering' |
           field == 'Education' |
           field == 'Humanities and arts')

```


```{r column graphs, echo=FALSE}

# Phds awarded by field over time (MALES)
males <- phds %>% 
  filter(sex == 'male') %>%
  group_by(year) %>% 
  mutate(prop = doctorates/sum(doctorates)*100)

field_levels <- c('Education',
                  'Humanities and arts',
                  'Engineering',
                  'Physical sciences and earth sciences')
males$field <- as_factor(males$field, levels = field_levels)
  
field_labs <- c('Physical and earth sciences',
                'Engineering',
                'Education',
                'Humanities and arts')

males_col <- ggplot(males, aes(x = year, y = prop,
                               fill = field)) +
  geom_col() +
  geom_text(aes(x = year, y = prop,
                label = doctorates),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_brewer(type = 'qual', palette = 3, direction = -1,
                    name = 'Field',
                    labels = field_labs) +
  labs(x = 'Year',
       y = 'Doctorates awarded (%)',
       title = 'Males') +
  theme_minimal()

males_col

# Phds awarded by field over time (FEMALES)
females <- phds %>% 
  filter(sex == 'female') %>%
  group_by(year) %>% 
  mutate(prop = doctorates/sum(doctorates)*100)
  
females$field <- as_factor(males$field, levels = field_levels)

females_col <- ggplot(females, aes(x = year, y = prop,
                                   fill = field)) +
  geom_col() +
  geom_text(aes(x = year, y = prop,
                label = doctorates),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_brewer(type = 'qual', palette = 3, direction = -1,
                    name = 'Field',
                    labels = field_labs) +
  theme_classic() +
  labs(x = 'Year',
       y = 'Doctorates awarded (%)',
       title = 'Females') +
  theme_minimal()

females_col

```


```{r chi squared, echo=FALSE, warning=FALSE}

# Make a contigency table of counts of doctorates per field over time (MALES ONLY)
males2 <- males %>% 
  select(field, year, doctorates) %>% 
  spread(key = year, value = doctorates)
males3 <- males2[,2:4]
rownames(males3) <- males2$field
males4 <- males2 %>% 
  mutate(prop_1985 = `1985`/sum(`1985`),
         prop_2000 = `2000`/sum(`2000`),
         prop_2015 = `2015`/sum(`2015`))

# Run chi-square test
male_chi <- chisq.test(males3)

male_chi

# Make a contigency table of counts of doctorates per field over time (FEMALES ONLY)
females2 <- females %>% 
  select(field, year, doctorates) %>% 
  spread(key = year, value = doctorates)
females3 <- females2[,2:4]
rownames(females3) <- females2$field
females4 <- females2 %>% 
  mutate(prop_1985 = `1985`/sum(`1985`),
         prop_2000 = `2000`/sum(`2000`),
         prop_2015 = `2015`/sum(`2015`))

# Run chi-square test
female_chi <- chisq.test(females3)

female_chi

```
