---
title: "esm-206-assignment-5-q4"
author: "Sean Denny"
date: "11/29/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 4:

**Exploring academic salaries for professors in U.S. colleges.** Explore relationships between variables in the ‘Faculty salary data (2008 - 2009 survey)’ dataset. Develop a model describing faculty salary based on data for faculty sex, rank, years in current position, field, and number of years since doctoral degree was earned. You should make decisions regarding which variables should remain in your final model. Describe the results qualitatively and quantitatively (i.e., don’t just report the statistical results of the model – make sure you describe interesting findings in text). You can also discuss any concerns that you have with the model(s) you present, if any.

```{r load packages and data, message = FALSE}
library(tidyverse)
library(stargazer)
prof_sal <- read_csv("Faculty salary data (2008 - 2009 survey).csv")
```

Rename discipline column:

```{r}
prof_sal <- prof_sal %>% 
  rename(discipline = "discipline_(A=theoretical_B=applied)")
```

### 1. Explore the data

```{r exploring by grouping/summarizing by explanatory variables, message = FALSE}

#rank

salary_rank <- prof_sal %>% 
  group_by(rank) %>% 
  summarize(
    mean = mean(salary)
)

#As expected.

#discipline; note: A = theoretical, B=applied

salary_discipline <- prof_sal %>% 
  group_by(discipline) %>% 
  summarize(
    mean = mean(salary)
  )

#Average of applied salaries is higher.

#years_since_phd

salary_years_phd <- prof_sal %>% 
  group_by(years_since_phd) %>% 
  summarize(
    mean = mean(salary)
  )

#Generally increases, but not consistent.

#years_faculty

salary_years_faculty <- prof_sal %>% 
  group_by(years_faculty) %>% 
  summarize(
    mean = mean(salary)
  )

#Appears to genereally increase, but not consistent.

#sex

salary_sex <- prof_sal %>% 
  group_by(sex) %>% 
  summarize(
    mean = mean(salary)
  )

#Average of males is higher.

```

### 2. Test for corellations

1. Coerce character variables into factors then numeric (i.e. change the class of the character data)

2. Test for correlations among all of the explanatory variables. Can we do this if the variables are factors, or do we absolutely have to change them to numeric? **Remember to create a new dataset without the response variable (salary) so you can run pairs() or cor() for the explanatory variables only.**

```{r testing for correlations}

cor_data <- select(prof_sal, -salary) %>% 
  mutate(rank = as.factor(rank)) %>% 
  mutate(rank = as.numeric(rank)) %>% 
  mutate(discipline = as.factor(discipline)) %>% 
  mutate(discipline = as.numeric(discipline)-1) %>% 
  mutate(sex = as.factor(sex)) %>% 
  mutate(sex = as.numeric(sex)-1)

cor(cor_data)
pairs(cor_data)

#Correlation for years_since_phd and rank = 0.53
#years_since_phd and years_faculty = 0.91

```

Before running the model, re-level the levels in rank so that Assistant Professor is the reference level. This isn't necessary but slightly easier for interpretation.

```{r}
prof_sal <- prof_sal %>% 
  mutate(rank = as.factor(rank))

prof_sal$rank <- fct_relevel(prof_sal$rank, "AsstProf")
```

### 3. Build the model

We're going to remove years_since_phd, which removes correlations with both years_faculty and rank. Also, these two variables (the ones we're keeping) are more interesting.

Use lm() with the following syntax: lm(y ~ x1 + x2 + x3..., data = df_name)

```{r}
prof_lm_1 <- lm(salary ~ sex + discipline + rank + years_faculty, data = prof_sal)

summary(prof_lm_1)
```

**Adjusted R-squared is 0.4407.** Meaning the variance in salary is not particularly well explained by variance in the... model output? explanatory variables?

```{r model diagnostics}
par(mfrow = c(2,2))
plot(prof_lm_1)
```

**Problems with homescedasticity. The data appear heteroscedastic.** *Also, data isn't normally distributed at upper values?*

**Is there a need to try other models? The variables we included are all interesting and make sense to include.** No, it doesn't.

**Remember that years_since_phd still is an explanatory factor.**

### 4. Figures

```{r stargazer table with model outputs}

#Prepare a regression table:

stargazer(prof_lm_1)

# If you want to work with this in Word, save to html, open, copy and paste into Word. Note that you specify the output 'type' as html here:

#stargazer(prof_lm_1, type = "html")

#stargazer(home_lm, home_lm2, type= "html",
         # title="My great title",
         # digits = 2) #digits=2 rounds the results to two digits


```

