---
title: "esm-206-assignment-5-q1"
author: "Sean Denny"
date: "11/29/2018"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Q1: Male and female graduate enrollment (1967 - 2015)

Compare trends in total graduate enrollment for males and females (including full-time/part-time and private/public universities) in the United States from 1967 - 2015. Describe your results statistically, graphically and in text.

```{r packages & data, include=FALSE}

# Load necessary packages

library(tidyverse)
library(stargazer)

# Load necessary data

enroll <- read_csv('Grad enrollment 1967 - 2015.csv',
                   na = '---') # total postbaccalaureate fall enrollment in degree-granting postsecondary institutions, by attendance status, sex of student, and control of institution: 1967 through 2015

```


```{r reformatting dataframes, include=FALSE}

# Dataframe for male vs. female, fulltime vs parttime enrollment
enroll_full <- enroll %>% 
  select(year, full_m, full_f) %>% 
  rename(male = full_m,
         female = full_f) %>% 
  gather(key = sex, value = enrollment, 2:3) %>% 
  mutate(status = 'fulltime')

enroll_part <- enroll %>% 
  select(year, part_m, part_f) %>% 
  rename(male = part_m,
         female = part_f) %>% 
  gather(key = sex, value = enrollment, 2:3) %>% 
  mutate(status = 'parttime')

enrollment <- rbind(enroll_full, enroll_part) %>% 
  select(year, sex, status, enrollment)

# Dataframe for public vs. private enrollment
enroll_public <- enroll %>%
  select(year, public) %>% 
  rename(enrollment = public) %>% 
  mutate(sector = 'public',
         type = NA)

enroll_private <- enroll %>% 
  select(year, nonprofit, forprofit) %>% 
  gather(key = type, value = enrollment, 2:3) %>% 
  mutate(sector = 'private')

enroll_sector <- rbind(enroll_public, enroll_private) %>% 
  select(year, sector, type, enrollment)

```


```{r scatter plots, echo=FALSE, warning=FALSE}

# Scatter plot: totals
total_scatter <- enrollment %>% 
  group_by(year, sex) %>% 
  summarize(total = sum(enrollment)) %>% 
  ggplot(aes(x = year, y = total)) +
    geom_point(aes(color = sex)) +
    theme_classic() +
    labs(x = 'Year',
         y = 'Enrollment (number of students)',
         title = 'Total male vs. female enrollment (1967-2015)')

total_scatter

# Scatter plot: male vs female, fulltime vs parttime
enroll_scatter <- ggplot(enrollment, aes(x = year,
                                         y = enrollment/1000)) +
  geom_point(aes(color = sex, shape = status)) +
  scale_shape_manual(values = c(19, 1),
                     name = 'Enrollment Status',
                     labels = c('Full-time', 'Part-time')) +
  scale_color_manual(values = c('orange', 'blue'),
                     name = 'Sex',
                     labels = c('Female', 'Male')) +
  labs(x = 'Year',
       y = 'Enrollment (thousands of students)',
       title = 'Enrollment over time (1967-2015)') +
  scale_y_continuous(limits = c(0, 1000)) +
  theme_classic()
enroll_scatter

# Scatter plot: public vs private
sector_scatter <- enroll_sector %>% 
  group_by(year, sector) %>% 
  summarize(total = sum(enrollment)) %>% 
  ggplot(aes(x = year, y = total)) +
    geom_point(aes(color = sector)) +
    scale_color_manual(values = c('grey60', 'black')) +
    theme_classic() +
    labs(x = 'Year',
         y = 'Enrollment (number of students)',
         title = 'Enrollment, public vs. private sector (1967-2015)')

sector_scatter

```


```{r linear regression: total male enrollment, echo=FALSE}

# Linear regression for total male enrollment
totals_m <- enrollment %>% 
  group_by(year, sex) %>% 
  summarize(total = sum(enrollment)) %>% 
  filter(sex == 'male')

male_lm <- lm(total ~ year, data = totals_m)

summary(male_lm)

# Model diagnostics
par(mfrow = c(2,2))
plot(male_lm)

# Pearson's r
male_pr <- cor.test(totals_m$year, totals_m$total)
male_pr

```

On average, total male enrollment increased by 9069 students per year. Year was a sigificant predictor of total male enrollment (b = 9069, t(47) = 16.6, p < 0.001), with a strong positive correlation between the two (Pearson's r = 0.92). The overall model (enrollment = 9096 * year -17112153) explains a siginificant amount of variance in enrollment over time (F(1,47) = 276, p < 0.001, R^2^ = 0.85).

```{r linear regression: total female enrollment, echo=FALSE}

# Linear regression for total female enrollment
totals_f <- enrollment %>% 
  group_by(year, sex) %>% 
  summarize(total = sum(enrollment)) %>% 
  filter(sex == 'female')

female_lm <- lm(total ~ year, data = totals_f)

summary(female_lm)

# Model diagnostics
par(mfrow = c(2,2))
plot(female_lm)

# Pearson's r
female_pr <- cor.test(totals_f$year, totals_f$total)
female_pr

```

On average, total female enrollment increased by 30,126 students per year. Year was a sigificant predictor of total female enrollment (b = 30126, t(47) = 51.6, p < 0.001), with a strong correlation between the two (Pearson's r = 0.99). The overall model (enrollment = 30126 * year -58955502) explains a siginificant amount of variance in enrollment over time (F(1,47) = 2669, p < 0.001, R^2^ = 0.98).

```{r linear regression: male vs female and full vs part, include=FALSE}

### MALE - FULLTIME ###

# Linear regression for male fulltime enrollment
totals_m_f <- enrollment %>% 
  mutate(enrollment = enrollment/1000) %>% 
  group_by(year, sex, status) %>% 
  summarize(total = sum(enrollment)) %>% 
  filter(sex == 'male',
         status == 'fulltime')

male_full_lm <- lm(total ~ year, data = totals_m_f)

summary(male_full_lm)

# Model diagnostics
par(mfrow = c(2,2))
plot(male_full_lm)

# Pearson's r
male_full_pr <- cor.test(totals_m_f$year, totals_m_f$total)
male_full_pr

### FEMALE - FULLTIME ###

# Linear regression for female fulltime enrollment
totals_f_f <- enrollment %>% 
  mutate(enrollment = enrollment/1000) %>% 
  group_by(year, sex, status) %>% 
  summarize(total = sum(enrollment)) %>% 
  filter(sex == 'female',
         status == 'fulltime')

female_full_lm <- lm(total ~ year, data = totals_f_f)

summary(female_full_lm)

# Model diagnostics
par(mfrow = c(2,2))
plot(female_full_lm)

# Pearson's r
female_full_pr <- cor.test(totals_f_f$year, totals_f_f$total)
female_full_pr

### MALE - PARTTIME ###

# Linear regression for male fulltime enrollment
totals_m_p <- enrollment %>%
  mutate(enrollment = enrollment/1000) %>% 
  group_by(year, sex, status) %>% 
  summarize(total = sum(enrollment)) %>% 
  filter(sex == 'male',
         status == 'parttime')

male_part_lm <- lm(total ~ year, data = totals_m_p)

summary(male_part_lm)

# Model diagnostics
par(mfrow = c(2,2))
plot(male_part_lm)

# Pearson's r
male_part_pr <- cor.test(totals_m_p$year, totals_m_p$total)
male_part_pr

### FEMALE - PARTTIME ###

# Linear regression for female fulltime enrollment
totals_f_p <- enrollment %>% 
  mutate(enrollment = enrollment/1000) %>% 
  group_by(year, sex, status) %>% 
  summarize(total = sum(enrollment)) %>% 
  filter(sex == 'female',
         status == 'parttime')

female_part_lm <- lm(total ~ year, data = totals_f_p)

summary(female_part_lm)

# Model diagnostics
par(mfrow = c(2,2))
plot(female_part_lm)

# Pearson's r
female_part_pr <- cor.test(totals_f_p$year, totals_f_p$total)
female_part_pr

```


```{r graph of lms, echo=FALSE}

# Plot: linear regressions for male vs female, fulltime vs parttime
enroll_lms <- ggplot(totals_m_f, aes(x = year, y = total/1000)) +
  geom_smooth(method = 'lm',
              color = 'blue', fill = 'blue',
              show.legend = TRUE) +
  geom_smooth(data = totals_m_p,
              aes(x = year, y = total/1000),
              method = 'lm',
              color = 'blue', fill = 'blue', linetype = 'dashed',
              show.legend = TRUE) +
  geom_smooth(data = totals_f_f,
              aes(x = year, y = total/1000),
              method = 'lm',
              color = 'orange', fill = 'orange',
              show.legend = TRUE) +
  geom_smooth(data = totals_f_p,
              aes(x = year, y = total/1000),
              method = 'lm',
              color = 'orange', fill = 'orange', linetype = 'dashed',
              show.legend = TRUE) +
  labs(x = 'Year',
       y = 'Enrollment (thousands of students)',
       title = 'Enrollment over time (1967-2015)') +
  scale_y_continuous(limits = c(0, 1000)) +
  theme_classic()

enroll_lms

# Scatter plot
enroll_scatter <- ggplot(enrollment, aes(x = year,
                                         y = enrollment/1000)) +
  geom_point(aes(color = sex, shape = status)) +
  scale_shape_manual(values = c(19, 1),
                     name = 'Enrollment Status',
                     labels = c('Full-time', 'Part-time')) +
  scale_color_manual(values = c('orange', 'blue'),
                     name = 'Sex',
                     labels = c('Female', 'Male')) +
  labs(x = 'Year',
       y = 'Enrollment (thousands of students)',
       title = 'Enrollment over time (1967-2015)') +
  scale_y_continuous(limits = c(0, 1000)) +
  theme_classic()

enroll_scatter

# Lines for legend
v1 <- c(1,2,3,4,5,6,7,8,9,10)
v2 <- c(1,2,1,1,1,1,1,1,2,1)
v3 <- c(2,3,2,2,2,2,2,2,3,2)
v4 <- c(3,4,3,3,3,3,3,3,4,3)
v5 <- c(4,5,4,4,4,4,4,4,5,4)
df <- data_frame(v1,v2,v3,v4,v5)
plot <- ggplot(df, aes(x = v1, v2)) +
  geom_smooth(method = 'lm',
              color = 'blue', fill = 'blue') +
  geom_smooth(aes(x = v1, y = v3),
              method = 'lm',
              color = 'blue', fill = 'blue', linetype = 'dashed') +
  geom_smooth(aes(x = v1, y = v4),
              method = 'lm',
              color = 'orange', fill = 'orange') +
  geom_smooth(aes(x = v1, y = v5),
              method = 'lm',
              color = 'orange', fill = 'orange', linetype = 'dashed') +
  theme_classic()
plot

```


```{r table of linear regression parameters, echo=FALSE, results='asis'}

# Table of linear regression parameters for male vs female and fulltime vs parttime models
stargazer(male_lm, female_lm,
          male_full_lm, male_part_lm,
          female_full_lm, female_part_lm,
          type = 'html',
          omit.stat = c('LL', 'ser'),
          covariate.labels = 'Year',
          dep.var.labels = 'Enrollment',
          column.labels = c('Total Male', 'Total Female',
                            'Male/Full-time', 'Male/Part-time',
                            'Female/Full-time', 'Female/Part-time'))

```

